\documentclass{seal_thesis}

\thesisType{Master Thesis}
\date{\today}
\title{AmbientTeams}
\subtitle{Staying socially connected in remote \\ knowledge work teams}
\author{Dario Bugmann}
\home{Aarau} % Geburtsort
\country{Switzerland}
\legi{15-708-852}
\prof{Prof. Dr. Thomas Fritz}
\assistent{Dr. André Meyer, Alexander Lill}
\email{dario.bugmann@uzh.ch}
\url{https://github.com/HASEL-UZH/PA.AmbientTeam}
\begindate{15.2.2021}
\enddate{22.7.2021}

\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage{tabularx}
\renewcommand{\mkbegdispquote}[2]{\itshape}
\usepackage[toc,page]{appendix}
\usepackage[style=alphabetic]{biblatex}
\addbibresource{thesis.bib}


\usepackage{array}
\usepackage{arydshln}
\setlength\dashlinedash{0.2pt}
\setlength\dashlinegap{1.5pt}
\setlength\arrayrulewidth{0.3pt}

\renewcommand{\arraystretch}{1.1}

\begin{document}
\maketitle

\frontmatter

\begin{acknowledgements}
    I thank Prof. Dr. Thomas Fritz, Dr. André Meyer, and Alexander Lill for the opportunity to write this thesis, the time they took for it, and the constructive feedback they were always able to give me. I appreciate that they always listened to my thoughts and suggestions. I am very grateful for the beautiful trip to Vitznau where I could meet all the members of the HASE research group. A big thank you to all my study participants who took the time to help me with this project and gave such great feedback. I greatly appreciate my family and friends and am grateful for their support. Last but not least, a special thank you to Nora for all her love and support and the faith she always had in me.
\end{acknowledgements}

\begin{abstract}
    As remote work becomes more prevalent, informal and spontaneous conversations become less frequent because knowledge workers lack essential cues about their colleagues, such as their state of attention or current location.  The lack of such social interactions can lead to feelings of isolation at work. While existing approaches focus mainly on improving team awareness to ameliorate coordination and collaboration problems caused by remote work, fewer tools focus on fostering informal, spontaneous communication. To address this gap, our approach focuses on people, their moods and status, and opportunities for spontaneous interactions to create more social awareness. To this end, we developed AmbientTeams, an unobtrusive and informal tool that aims to reduce the perceived distance between distant colleagues. AmbientTeams strives to achieve this goal following a mood-based micro-blogging approach that allows individuals to share moods and status updates with the team and provides various opportunities for a subsequent response. In a preliminary evaluation, we tested our research prototype on a group of five knowledge workers who used the tool for one week. The results show that AmbientTeams facilitated getting to know each other by providing mood-based information. All in all, the encouraging results show that our novel approach, which allows knowledge workers to quickly and easily share moods with their team, has future potential as it enables and promotes a more informal, lightweight way of communicating.
\end{abstract}

\begin{zusammenfassung}
    Mit der zunehmenden Verbreitung von Fernarbeit werden informelle und spontane Gespräche seltener, da Wissensarbeiter keine wesentlichen Hinweise über ihre Kollegen erhalten, wie z. B. deren Aufmerksamkeitszustand oder aktuellen Standort.  Der Mangel an solchen sozialen Interaktionen kann zu Gefühlen der Isolation bei der Arbeit führen. Während sich bestehende Ansätze hauptsächlich auf die Verbesserung des Team-Bewusstseins konzentrieren, um Koordinations- und Kooperationsprobleme, die durch Fernarbeit verursacht werden, zu verbessern, konzentrieren sich weniger Tools auf die Förderung informeller, spontaner Kommunikation. Um diese Lücke zu schließen, konzentriert sich unser Ansatz auf Menschen, ihre Stimmungen und ihren Status sowie auf Gelegenheiten für spontane Interaktionen, um mehr soziales Bewusstsein zu schaffen. Zu diesem Zweck haben wir AmbientTeams entwickelt, ein unaufdringliches und informelles Tool, das darauf abzielt, die wahrgenommene Distanz zwischen entfernten Kollegen zu verringern. AmbientTeams strebt dieses Ziel an, indem es einen stimmungsbasierten Micro-Blogging-Ansatz verfolgt, der es Personen ermöglicht, Stimmungen und Status-Updates mit dem Team zu teilen und verschiedene Möglichkeiten für eine darauffolgende Reaktion bietet. In einer vorläufigen Evaluation haben wir unseren Forschungsprototyp an einer Gruppe von fünf Wissensarbeitern getestet, die das Tool eine Woche lang nutzten. Die Ergebnisse zeigen, dass AmbientTeams das gegenseitige Kennenlernen durch die Bereitstellung von stimmungsbasierten Informationen erleichtert hat. Alles in allem zeigen die ermutigenden Ergebnisse, dass unser neuartiger Ansatz, der es Wissensarbeitern ermöglicht, schnell und einfach Stimmungen mit ihrem Team zu teilen, Zukunftspotenzial hat, da er eine informellere, leichtere Art der Kommunikation ermöglicht und fördert.
\end{zusammenfassung}

\setcounter{tocdepth}{1}
\tableofcontents
\listoffigures
\listoftables
% \lstlistoflistings

\mainmatter
\include{chapters/introduction}
\include{chapters/related_work}
\include{chapters/approach}
\include{chapters/prototype}
\include{chapters/preliminary_evaluation}
\include{chapters/results}
\include{chapters/discussion}
\include{chapters/conclusion}

\begin{appendices}
    \include{chapters/additional_documents}
\end{appendices}

\backmatter
\printbibliography

\end{document}
